---
title: "City Factsheets"
author: "Thiyaghessan"
date: "2025-07-25"
output: 
  word_document:
    reference_docx: "C:/Users/thiya/Documents/Urban/NCCS/nonprofit_govt_grants/templates/Urban_Two-Page_Fact_Sheet_Template.dotm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
getwd()
city_summaries <- readxl::read_xlsx("C:/Users/thiya/Documents/Urban/NCCS/nonprofit_govt_grants/data/data_requests/summarized_metrics_intermediat.xlsx", sheet = 2)
```

# City Summaries

Across the nineteen cities, 12,747 public charities reported receiving a total 
of over $55 billion from government grants in 2021. The inflation adjusted total 
of over $62 billion is almost two times the most recent estimates of [corporate 
giving](https://philanthropy.indianapolis.iu.edu/news-events/news/_news/2024/giving-usa-us-charitable-giving-totaled-557.16-billion-in-2023.html).

In every city, the typical nonprofit could not cover its expenses without
government grants. And, between 60 and 80 percent of nonprofits that receive 
government grants would be at risk of a financial shortfall.

The table allows users to understand the financial risk to nonprofits across these cities
if they lose their government grants.

```{r, echo=FALSE}
rename_list <- list(
    "Number of 990 filers with government grants" = "num_990filers_govgrants",
    "Total government grants ($)" = "total_govt_grants",
    "Operating surplus with government grants (%)" = "median_profit_margin",
    "Operating surplus without government grants (%)" = "median_profit_margin_no_govt_grants",
    "Share of 990 filers with government grants at risk" = "proportion_at_risk"
  )
city_summaries <- city_summaries |>
  dplyr::select(! CENSUS_STATE_NAME) |>
  dplyr::rename("City" = CITY) |>
      dplyr::mutate(
        num_990filers_govgrants = scales::comma(num_990filers_govgrants),
        total_govt_grants = scales::dollar(total_govt_grants),
        median_profit_margin = scales::percent(median_profit_margin, accuracy = 0.01),
        median_profit_margin_no_govt_grants = scales::percent(median_profit_margin_no_govt_grants, accuracy = 0.01),
        proportion_at_risk = scales::percent(proportion_at_risk, accuracy = 0.01)
      ) |>
  dplyr::rename(!!!rename_list)
```

```{r}
city_summaries %>%
  kableExtra::kable()
```

