---
title: "Nonprofits, Financial Sustainability, and Government Grants in 2022"
author: 
subtitle: '<span style="font-size: 14pt; text-transform: capitalize;">Laura Tomasko,
  Jesse Lecy, Hannah Martin, Thiyaghessan Poongundranar</span>'
output:
  html_document:
    css: web_report.css
    self_contained: false
    theme: null
  pdf_document: default
params:
  geography: California
format: html
---

```{=html}
<!-- this calls Lato font --> 
<style>
@import url('https://fonts.googleapis.com/css?family=Lato&display=swap');
</style>
```

```{=html}
<!-- this calls crucial iframe resizing script --> 
<script type="text/javascript" src="https://apps.urban.org/features/dataviz-libs/pym-js/v1.3.2/pym.nprapps.org_pym.v1.min.js">
</script>

<script>
var pymChild = new pym.Child({ polling: 500 });
</script>
```

```{r rmarkdown-setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

```{r r-setup, echo=FALSE}
library(tidyverse)
library(urbnthemes)
library(gt)
library(gtExtras)

set_urbn_defaults(style = "print")
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r r-scripts, echo=FALSE}
source("R/build_gt_table.R")
```

This series of fact sheets provides three 2021 data points at the national and state level:

1.	**Number of 501(c)(3) public charities** – These data show the size of the charitable sector in each geography.

2.	**These 501(c)(3) public charities’ months of spending on hand** – Months of spending on hand measures nonprofits’ sustainability. It indicates the number of months a nonprofit could maintain its operational spending if it suddenly lost all its revenue. Experts recommend nonprofits maintain at least three months of spending on hand[^1]. 

    We use the following formula to calculate months of spending on hand:

    $$ 12 * \frac{\text{unrestricted net assets} - (\text{land, buildings, and equipment} - \text{tax-exempt bond liabilities} - \text{mortgages and notes payable})}{\text{total expenses} - \text{depreciation}} $$

    Using lines from the Form 990, this can formula can be written as follows:

    $$ 12 * \frac{\text{Part 10 line 27 EOY} - (\text{Part 10 line 10c EOY} - \text{Part 10 line 20 EOY} - \text{Part 10 line 23 EOY} - \text{Part 10 line 24 EOY})}{\text{Part 01 line 18 CY} - \text{Part 09 line 22 col A}} $$

    We have separated the number of months of spending on hand into two categories: with and without tangible assets. Tangible assets include land, buildings, and equipment.

3. **The aggregate dollar amount of government grants these 501(c)(3) public charities received** – These data point toward the importance of government grants to charities in each geography. 

4. **Distribution of Government Grant Revenue** - This metric shows what percentage of nonprofits in each area receive different levels of government grant funding, ranging from those highly dependent (80-100% of revenue) to those less reliant (0-20% of revenue) on government grants. The distribution is broken into five categories: 0-20%, 20-40%, 40-60%, 60-80%, and 80-100% of total revenue.

Each fact sheet displays these data points for the US or state overall and broken down by subsector and the size of their expenses.

<div class="under-table-spacing"></div>

# Analysis of `r params$geography`

<div class="over-table-spacing"></div>

#### By Size

```{r echo=FALSE}
build_gt_table("data/intermediate/factsheet_test_expense_df.csv", "EXPENSE_CATEGORY", "Size")
```

###### **Source**: *National Center for Charitable Statistics [2021 501(c)(3) Charities PZ Scope](https://urbaninstitute.github.io/nccs/catalogs/catalog-core.html), [2021 Efilers](https://nccs.urban.org/nccs/catalogs/catalog-efile.html) and [Unified  BMF](https://nccs.urban.org/nccs/datasets/bmf/)*

###### **Notes**: *Disclaimers about the data.*

<div class="under-table-spacing"></div>

The same data can be disaggregated by Subsector.

<div class="over-table-spacing"></div>

#### By Subsector

:::main_table
```{r echo=FALSE}
build_gt_table("data/intermediate/factsheet_test_subsector_df.csv", "SUBSECTOR", "Subsector")
```
:::


###### **Source**: *National Center for Charitable Statistics [2021 501(c)(3) Charities PZ Scope](https://urbaninstitute.github.io/nccs/catalogs/catalog-core.html), [2021 Efilers](https://nccs.urban.org/nccs/catalogs/catalog-efile.html) and [Unified  BMF](https://nccs.urban.org/nccs/datasets/bmf/)*

###### **Notes**: *Disclaimers about the data.*

<div class="under-table-spacing"></div>

[^1]: Bowman, Woods. 2011. “[Financial Capacity and Sustainability of Ordinary Nonprofits.](https://doi.org/10.1002/nml.20039)” Nonprofit Management & Leadership 22 (1): 37–51. 