---
title: ""
author: 
subtitle: ''
output:
  html_document:
    css: web_report.css
    self_contained: false
  pdf_document: default
params:
  geography: United States
---

```{=html}
<!-- this calls Lato font --> 
<style>
@import url('https://fonts.googleapis.com/css?family=Lato&display=swap');
</style>
```

```{=html}
<!-- this calls crucial iframe resizing script --> 
<script type="text/javascript" src="https://apps.urban.org/features/dataviz-libs/pym-js/v1.3.2/pym.nprapps.org_pym.v1.min.js">
</script>

<script>
var pymChild = new pym.Child({ polling: 500 });
</script>
```

```{r rmarkdown-setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

```{r r-setup, echo=FALSE}
library(tidyverse)
library(urbnthemes)
library(gt)
library(gtExtras)

set_urbn_defaults(style = "print")
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## `r params$geography` - Financial Risk of Nonprofits Losing Government Grants

<div class="over-table-spacing"></div>

```{r r-data, echo=FALSE}
source("R/format_percentages.R")

national_bystate <- data.table::fread("data/processed/national_bystate.csv")
national_bysize <- data.table::fread("data/processed/national_bysize.csv")
national_bysubsector <- data.table::fread("data/processed/national_bysubsector.csv")
```

```{r formatting, echo=FALSE}
national_bystate <- format_percentages(national_bystate, 
                             "Share of 990 filers with government grants at risk",
                             "formatted_percentages")
national_bysize <- format_percentages(national_bysize,
                           "Share of 990 filers with government grants at risk",
                           "formatted_percentages")
national_bysubsector <- format_percentages(national_bysubsector,
                                "Share of 990 filers with government grants at risk",
                                "formatted_percentages")
```

In the United States, nonprofits received a total of $267B in government grants. Nonprofits in California, New York, and the District of Columbia received the most government funding. Nonprofits in Alaska, West Virginia, and Louisiana would have the biggest gap to fill if they lost their government grants. Nonprofits in Alaska, West Virginia, and Arkansas have the highest share of nonprofits at risk of not covering their expenses if they lost their government grants.

## {.tabset}

### State
```{r}
national_bystate  |> mutate(State = factor(
  State,
  levels = c(
    "United States",
    "Alabama",
    "Alaska",
    "Arizona",
    "Arkansas",
    "California",
    "Colorado",
    "Connecticut",
    "Delaware",
    "District of Columbia",
    "Florida",
    "Georgia",
    "Hawaii",
    "Idaho",
    "Illinois",
    "Indiana",
    "Iowa",
    "Kansas",
    "Kentucky",
    "Louisiana",
    "Maine",
    "Maryland",
    "Massachusetts",
    "Michigan",
    "Minnesota",
    "Mississippi",
    "Missouri",
    "Montana",
    "Nebraska",
    "Nevada",
    "New Hampshire",
    "New Jersey",
    "New Mexico",
    "New York",
    "North Carolina",
    "North Dakota",
    "Ohio",
    "Oklahoma",
    "Oregon",
    "Other U.S. Territories",
    "Pennsylvania",
    "Rhode Island",
    "South Carolina",
    "South Dakota",
    "Tennessee",
    "Texas",
    "Utah",
    "Vermont",
    "Virginia",
    "Washington",
    "West Virginia",
    "Wisconsin",
    "Wyoming",
    "Other US Jurisdictions/Unmapped"
  )
)) |>
  arrange(State) |>
  dplyr::mutate(color = "") |>
  gt() |>
  data_color(
    columns = formatted_percentages,
    target_columns = color,
    method = "auto",
    palette = c(
      "#FFF2CF",
      "#FCE39E",
      "#FDD870",
      "#FCCB41",
      "#FDBF11",
      "#E88E2D",
      "#CA5800",
      "#843215"
    )
  ) |>
  cols_hide(formatted_percentages) |>
  cols_width(color ~ px(5)) |>
  cols_label(color = "") |>
  cols_align(align = "left", columns = c("State")) |>
  tab_style(
    style = cell_borders(
      sides = "right",
      color = "#d2d2d2",
      weight = px(1)
    ),
    locations = cells_body(columns = 1)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_body(columns = 1), cells_column_labels())
  ) |>
  tab_options(table.border.top.style = "hidden",
              data_row.padding = px(10))
```

### Size
```{r}
national_bysize |> mutate(Size = factor(
  Size,
  levels = c(
    "Less than $100K",
    "$100K to $499K",
    "$500K to $999K",
    "$1M to $4.9M",
    "$5M to $9.9M",
    "$10M or more",
    "Total"
  )
)) |>
  arrange(Size) |>
   dplyr::mutate(color = "") |>
  gt() |>
  data_color(
    columns = formatted_percentages,
    target_columns = color,
    method = "auto",
    palette = c("#FFF2CF", "#FCE39E", "#FDD870", "#FCCB41", "#FDBF11", "#E88E2D", "#CA5800", "#843215")
    ) |>
  cols_hide(formatted_percentages) |>
    cols_width(color ~ px(5)) |>
  cols_label(
    color = ""  # This sets the column name to an empty string
) |>
  cols_align(
    align = "left",
    columns = c("Size")
  ) |>
  tab_style(
    style = cell_borders(
      sides = "right",
      color = "#d2d2d2",
      weight = px(1)
    ),
    locations = cells_body(
      columns = 1
    )
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(
      cells_body(columns = 1),  # First column
      cells_column_labels()      # All headers
    )
  ) |>
  tab_options(
    table.border.top.style = "hidden",
    data_row.padding = px(10)
  )
```

### Subsector
```{r}
national_bysubsector |> mutate(Subsector = factor(
  Subsector,
  levels = c(
        "Arts, culture, and humanities",
        "Education",
        "Environment and animals",
        "Health",
        "Hospitals",
        "Human services",
        "International, foreign affairs",
        "Public, societal benefit",
        "Religion-related",
        "Mutual/membership benefit",
        "Universities",
        "Unclassified",
        "Total"
      )
)) |>
  arrange(Subsector) |>
  dplyr::mutate(color = "") |>
  gt() |>
  data_color(
    columns = formatted_percentages,
    target_columns = color,
    method = "auto",
    palette = c(
      "#FFF2CF",
      "#FCE39E",
      "#FDD870",
      "#FCCB41",
      "#FDBF11",
      "#E88E2D",
      "#CA5800",
      "#843215"
    )
  ) |>
  cols_hide(formatted_percentages) |>
  cols_width(color ~ px(5)) |>
  cols_label(color = "") |>
  cols_align(align = "left", columns = c("Subsector")) |>
  tab_style(
    style = cell_borders(
      sides = "right",
      color = "#d2d2d2",
      weight = px(1)
    ),
    locations = cells_body(columns = 1)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = list(cells_body(columns = 1), cells_column_labels())
  ) |>
  tab_options(table.border.top.style = "hidden",
              data_row.padding = px(10))
                
```

## {-}

###### **Notes**: Operating surplus reported as the median.

###### **Source**: National Center for Charitable Statistics 2021 501(c)(3) Charities [E-file Data](https://nccs.urban.org/nccs/datasets/efile/).