---
title: ""
author:
subtitle: ''
output:
  html_document:
    css: web_report.css
    self_contained: false
  pdf_document: default
params:
  geography: United States
  year: 2021
---

```{=html}
<!-- this calls Lato font -->
<style>
@import url('https://fonts.googleapis.com/css?family=Lato&display=swap');
</style>
```

```{=html}
<!-- this calls crucial iframe resizing script -->
<script type="text/javascript" src="https://apps.urban.org/features/dataviz-libs/pym-js/v1.3.2/pym.nprapps.org_pym.v1.min.js">
</script>

<script>
var pymChild = new pym.Child({ polling: 500 });
</script>
```

```{r rmarkdown-setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

```{r r-setup, echo=FALSE}
library(tidyverse)
library(urbnthemes)
library(gt)
library(gtExtras)
library(epoxy)

set_urbn_defaults(style = "print")
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## `r params$geography` - How Losing Government Grants Would Have Affected Nonprofit Finances in `r params$year`

<div class="over-table-spacing"></div>

```{r r-data, echo=FALSE}
setwd(rprojroot::find_rstudio_root_file())
source("R/config.R")
source("R/format_percentages.R")
source("R/format_gt_table.R")

data_dir <- dir_processed_year(params$year)
national_bystate <- data.table::fread(file.path(data_dir, "national_bystate.csv"))
national_bysize <- data.table::fread(file.path(data_dir, "national_bysize.csv"))
national_bysubsector <- data.table::fread(file.path(data_dir, "national_bysubsector.csv"))

total_filers <- format(
  sum(as.numeric(gsub(",", "", national_bystate$`Number of 990 filers with government grants`[
    national_bystate$State == "United States"
  ]))),
  big.mark = ","
)
```

```{r formatting, echo=FALSE}
national_bystate <- format_percentages(national_bystate,
                             "Share of 990 filers with government grants at risk",
                             "formatted_percentages")
national_bysize <- format_percentages(national_bysize,
                           "Share of 990 filers with government grants at risk",
                           "formatted_percentages")
national_bysubsector <- format_percentages(national_bysubsector,
                                "Share of 990 filers with government grants at risk",
                                "formatted_percentages")
```

```{r intro-stats, echo=FALSE}
# Total government grants (national)
total_govgrants <- national_bystate |>
  dplyr::filter(State == "United States") |>
  dplyr::pull(`Total government grants ($)`)
total_govgrants <- as.numeric(gsub("[$,]", "", total_govgrants))
total_govgrants <- scales::label_number(accuracy = 0.1, scale_cut = scales::cut_short_scale())(total_govgrants)

# Top 3 states by total government grants
topstates_grants <- national_bystate |>
  dplyr::filter(State != "United States",
                State != "Other U.S. Territories",
                State != "Other/unmapped jurisdictions") |>
  dplyr::mutate(val = as.numeric(gsub("[$,]", "", `Total government grants ($)`))) |>
  dplyr::arrange(desc(val)) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(State)

# Top 3 states by biggest gap (most negative surplus without govt grants)
topstates_surplus <- national_bystate |>
  dplyr::filter(State != "United States",
                State != "Other U.S. Territories",
                State != "Other/unmapped jurisdictions") |>
  dplyr::mutate(val = as.numeric(gsub("[%]", "", `Size of operating surplus without government grants`))) |>
  dplyr::arrange(val) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(State)

# Top 3 states by share at risk
topstates_atrisk <- national_bystate |>
  dplyr::filter(State != "United States",
                State != "Other U.S. Territories",
                State != "Other/unmapped jurisdictions") |>
  dplyr::mutate(val = as.numeric(gsub("[%]", "", `Share of 990 filers with government grants at risk`))) |>
  dplyr::arrange(desc(val)) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(State)

# Format as "X, Y, and Z"
fmt_list <- function(x) gsub(",([^,]*)$", ", and\\1", paste(x, collapse = ", "))
topstates_grants <- fmt_list(topstates_grants)
topstates_surplus <- fmt_list(topstates_surplus)
topstates_atrisk <- fmt_list(topstates_atrisk)

intro_text <- glue::glue(
  "In the United States, nonprofits received a total of ${total_govgrants} ",
  "in government grants. Nonprofits in {topstates_grants} received the most ",
  "government funding. Nonprofits in {topstates_surplus} would have the biggest ",
  "gap to fill if they lost their government grants. Nonprofits in ",
  "{topstates_atrisk} have the highest share of nonprofits at risk of not ",
  "covering their expenses if they lost their government grants."
)
```

```{epoxy}
{intro_text}
```

## {.tabset}

### State
```{r}
national_bystate |>
  mutate(State = factor(State, levels = STATE_NAME_LEVELS)) |>
  arrange(State) |>
  gt() |>
  style_factsheet_table()
```

### Size
```{r}
national_bysize |>
  mutate(Size = factor(Size, levels = c(EXPENSE_LABELS, "Total"))) |>
  arrange(Size) |>
  dplyr::mutate(color = "") |>
  gt() |>
  cols_width(color ~ px(5)) |>
  cols_label(color = "") |>
  style_factsheet_table()
```

### Subsector
```{r}
national_bysubsector |>
  mutate(Subsector = factor(Subsector, levels = c(SUBSECTOR_LEVELS, "Total"))) |>
  arrange(Subsector) |>
  dplyr::mutate(color = "") |>
  gt() |>
  cols_width(color ~ px(5)) |>
  cols_label(color = "") |>
  style_factsheet_table()
```

## {-}

###### **Notes**: Includes `r total_filers` nonprofits that electronically filed an IRS Form 990 and reported receiving government grants in tax year `r params$year`. Government grants are from governments at the local, state, federal, and territory levels, as well as foreign governments. A positive operating surplus indicates a nonprofit can cover all of its expenses and has funds available to reinvest in its mission. A negative surplus indicates the nonprofit could not do so in that tax year. Operating surplus reported as the median.

###### **Source**: National Center for Charitable Statistics `r params$year` 501(c)(3) Charities [E-file Data](https://nccs.urban.org/nccs/datasets/efile/).
