---
title: ""
author:
subtitle: ''
output:
  html_document:
    css: web_report.css
    self_contained: false
  pdf_document: default
params:
  state: California
  year: 2021
---

```{=html}
<!-- this calls Lato font -->
<style>
@import url('https://fonts.googleapis.com/css?family=Lato&display=swap');
</style>
```

```{=html}
<!-- this calls crucial iframe resizing script -->
<script type="text/javascript" src="https://apps.urban.org/features/dataviz-libs/pym-js/v1.3.2/pym.nprapps.org_pym.v1.min.js">
</script>

<script>
var pymChild = new pym.Child({ polling: 500 });
</script>
```

```{r rmarkdown-setup, echo=FALSE}
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(echo = FALSE)
```

```{r r-setup, echo=FALSE}
library(tidyverse)
library(urbnthemes)
library(gt)
library(gtExtras)
library(stringr)
library(epoxy)

set_urbn_defaults(style = "print")
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## `r params$state` - How Losing Government Grants Would Affect Nonprofit Finances

<div class="over-table-spacing"></div>


```{r load-data, echo=FALSE}
setwd(rprojroot::find_rstudio_root_file())
source("R/config.R")
source("R/format_percentages.R")
source("R/format_districts.R")
source("R/format_gt_table.R")

state <- gsub(" ", "-", tolower(params$state))
factsheet_dir <- dir_state_factsheets_year(params$year)

county <- data.table::fread(sprintf(file.path(factsheet_dir, "%s_bycounty.csv"),
                                    state))
district <- data.table::fread(sprintf(file.path(factsheet_dir, "%s_bydistrict.csv"),
                                      state))
size <- data.table::fread(sprintf(file.path(factsheet_dir, "%s_bysize.csv"),
                                  state))
subsector <- data.table::fread(sprintf(file.path(factsheet_dir, "%s_bysubsector.csv"),
                                       state))

total_filers <- format(
  sum(as.numeric(gsub(",", "", size$`Number of 990 filers with government grants`[
    size$Size == "Total"
  ]))),
  big.mark = ","
)
```

```{r into-para, echo=FALSE}
state_title <- params$state
total_govgrants <- size |>
  dplyr::filter(Size == "Total") |>
  dplyr::pull(`Total government grants ($)`)

total_govgrants <- gsub(
  "[$,]", "", total_govgrants)
total_govgrants <- as.numeric(total_govgrants)
total_govgrants <- scales::label_number(accuracy=0.1, scale_cut=scales::cut_short_scale())(total_govgrants)

topcounties_grants <- county |>
  dplyr::filter(Geography != params$state, Geography != "United States") |>
  dplyr::mutate(govfunding = as.numeric(gsub("[$,]", "", `Total government grants ($)`))) |>
  dplyr::arrange(desc(govfunding)) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(Geography)

topdistricts_grants <- district |>
  dplyr::filter(Geography != params$state, Geography != "United States") |>
  dplyr::mutate(govfunding = as.numeric(gsub("[$,]", "", `Total government grants ($)`))) |>
  dplyr::arrange(desc(govfunding)) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(Geography)

topcounties_surplus <- county |>
  dplyr::filter(Geography != params$state, Geography != "United States") |>
  dplyr::mutate(govfunding = as.numeric(
    gsub("[%]", "", `Size of operating surplus without government grants`)
  )) |>
  dplyr::arrange(govfunding) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(Geography)

topdistricts_surplus <- district |>
  dplyr::filter(Geography != params$state, Geography != "United States") |>
  dplyr::mutate(govfunding = as.numeric(
    gsub("[%]", "", `Size of operating surplus without government grants`)
  )) |>
  dplyr::arrange(govfunding) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(Geography)

topcounties_atrisk <- county |>
  dplyr::filter(Geography != params$state, Geography != "United States") |>
  dplyr::mutate(govfunding = as.numeric(
    gsub("[%]", "", `Share of 990 filers with government grants at risk`)
  )) |>
  dplyr::arrange(desc(govfunding)) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(Geography)

topdistricts_atrisk <- district |>
  dplyr::filter(Geography != params$state, Geography != "United States") |>
  dplyr::mutate(govfunding = as.numeric(
    gsub("[%]", "", `Share of 990 filers with government grants at risk`)
  )) |>
  dplyr::arrange(desc(govfunding)) |>
  dplyr::slice_head(n = 3) |>
  dplyr::pull(Geography)

topcounties_grants <- paste(topcounties_grants, collapse = ", ")
topcounties_surplus <- paste(topcounties_surplus, collapse = ", ")
topcounties_atrisk <- paste(topcounties_atrisk, collapse = ", ")

topcounties_grants <- gsub(",([^,]*)$", ", and\\1", topcounties_grants)
topcounties_grants <- gsub(" County", "", topcounties_grants)
topcounties_surplus <- gsub(",([^,]*)$", ", and\\1", topcounties_surplus)
topcounties_surplus <- gsub(" County", "", topcounties_surplus)
topcounties_atrisk <- gsub(",([^,]*)$", ", and\\1", topcounties_atrisk)
topcounties_atrisk <- gsub(" County", "", topcounties_atrisk)

topdistricts_grants <- format_districts(topdistricts_grants)
topdistricts_surplus <- format_districts(topdistricts_surplus)
topdistricts_atrisk <- format_districts(topdistricts_atrisk)
```

```{r, echo=FALSE}
if (state_title != "District of Columbia"){
  text <- glue::glue(
    "In {state_title}, nonprofits received a total of ${total_govgrants} in government grants. Nonprofits in {topcounties_grants} counties {topdistricts_grants} received the most government funding. Nonprofits in {topcounties_surplus} counties {topdistricts_surplus} would have the biggest gap to fill if they lost their government grants. Nonprofits in {topcounties_atrisk} counties {topdistricts_atrisk} have the highest share of nonprofits at risk of not covering their expenses if they lost their government grants."
  )
} else {
  text <- glue::glue("In the District of Columbia, nonprofits received a total of ${total_govgrants} in government grants.")
}
```

```{epoxy}
{text}
```

```{r formatting, echo=FALSE}
county <- format_percentages(county,
                             "Share of 990 filers with government grants at risk",
                             "formatted_percentages")
district <- format_percentages(district,
                               "Share of 990 filers with government grants at risk",
                               "formatted_percentages")
size <- format_percentages(size,
                           "Share of 990 filers with government grants at risk",
                           "formatted_percentages")
subsector <- format_percentages(subsector,
                                "Share of 990 filers with government grants at risk",
                                "formatted_percentages")
```


## {.tabset}

### Counties
```{r}
county |>
  dplyr::distinct() |>
  dplyr::mutate(Geography = factor(Geography)) |>
  dplyr::mutate(color = "") |>
  gt() |>
  cols_width(color ~ px(5)) |>
  cols_label(color = "") |>
  style_factsheet_table()
```

### Congressional districts
```{r}
district |>
  dplyr::mutate(color = "") |>
  gt() |>
  cols_width(color ~ px(5)) |>
  cols_label(color = "") |>
  style_factsheet_table()
```

### Size
```{r}
size |>
  mutate(Size = factor(Size, levels = c(EXPENSE_LABELS, "Total"))) |>
  arrange(Size) |>
  dplyr::mutate(color = "") |>
  gt() |>
  cols_width(color ~ px(5)) |>
  cols_label(color = "") |>
  style_factsheet_table()
```

### Subsector
```{r}
subsector |>
  mutate(Subsector = factor(Subsector, levels = c(SUBSECTOR_LEVELS, "Total"))) |>
  arrange(Subsector) |>
  dplyr::mutate(color = "") |>
  gt() |>
  cols_width(color ~ px(5)) |>
  cols_label(color = "") |>
  style_factsheet_table()
```

## {-}

###### **Notes**: Includes `r total_filers` nonprofits that electronically filed an IRS Form 990 and reported receiving government grants in tax year `r params$year`. Government grants are from governments at the local, state, federal, and territory levels, as well as foreign governments. A positive operating surplus indicates a nonprofit can cover all of its expenses and has funds available to reinvest in its mission. A negative surplus indicates they could not do so in that tax year. Operating surplus reported as the median.

###### **Source**: National Center for Charitable Statistics `r params$year` 501(c)(3) Charities [E-file Data](https://nccs.urban.org/nccs/datasets/efile/).
